Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-12 manuscript/manuscript.tex
minor changes

which estimates the \vb{impact of selection} on allele frequency change between
the initial generation $0$ and some later generation $t$, which can be varied
to see how this quantity grows through time. When the sum of the covariances is
positive, this measure can intuitively be understood as the relative fraction
of allele frequency change normally thought of as ``drift" that is actually due
to selection. Additionally, $G(t)$ can be understood as a short-timescale
estimate of the reduction in neutral diversity due to linked selection (or
equivalently the reduction in neutral effective population size needed to
account linked selection, see Supplementary Materials Section
\ref{supp:reduction-approx}). Since \textcite{Barghi2019-qy} experiment is
sequenced every ten generations, in the numerator for the covariance we use the
allele frequency changes between adjacent timepoints, which are ten generations
apart; \vb{thus the strong covariances between adjacent generations are not
  observable and do not contribute to the numerator of $G(t)$}.
  \vb{Additionally, selection also inflates the variance in allele frequency
    change per generation; however, this effect cannot be easily distinguished
  from drift.} \vb{For both these reasons, our measure $G(t)$ is strongly
conservative} (\vb{we demonstrate this through simulations in Supplementary
Material Section \ref{supp:tempblocks}}).  Still, we find a remarkably strong
signal. Greater than $20\%$ of total, genome-wide allele frequency change over
60 generations is the result of selection (Figure \ref{fig:figure-1} B).

@@

We observed that in the longest study  we analyzed \parencite{Barghi2019-qy},
some genome-wide temporal covariances become negative at future timepoints (see
the first two rows in Figure \ref{fig:figure-1} A). This shows that alleles
that were on average going up initially are later going down in frequency, i.e.
that the average direction of selection experienced by alleles has flipped.
This must reflect either a change in the environment or the genetic background,
due to epistatic relationships among alleles altered by frequency changes or
recombination breaking up selective alleles.  Such reversals of selective
dynamics could be occurring at other timepoints but the signal of a change in
the direction of selection at particular loci may be washed out when we
calculate our genome-wide average temporal covariances. To address this
limitation, we calculated the distribution of the temporal covariances over
100kb windowed covariances (Figure \ref{fig:figure-3} shows these distributions
pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria, which can occur over long physical distances in
E\&R and selection studies \parencite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}. To
address this, we have developed a permutation-based procedure that constructs
an empirical \vb{neutral} null distribution by randomly flipping the sign of
the allele frequency changes in each genomic window (i.e. a single random sign
flip is applied to all loci in a window). This destroys the systematic
covariances created by linked selection and creates a sampling distribution of
the covariances spuriously created by neutral genetic drift while preserving
the complex dependencies between adjacent loci created by linkage
disequilibrium.  This empirical neutral null distribution is conservative in
the sense that the variances of the covariances are wider than expected under
drift alone as they include the effect of selection on the allele frequency
change within a time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that there are an empirical excess of windows with
positive covariances between close timepoints compared to the null distribution
(a heavier right tail), and that this then shifts to an excess of windows with
negative covariances between more distant timepoints (a heavier left tail)

>>>>

which estimates the \vb{impact of selection} on allele frequency change between
the initial generation $0$ and some later generation $t$, which can be varied
to see how this quantity grows through time. When the sum of the covariances is
positive, this measure can intuitively be understood as the relative fraction
of allele frequency change normally thought of as ``drift" that is actually due
to selection. Additionally, $G(t)$ can be understood as a short-timescale
estimate of the reduction in neutral diversity due to linked selection (or
equivalently the reduction in neutral effective population size needed to
account linked selection, see Supplementary Materials Section
\ref{supp:reduction-approx}). Since \textcite{Barghi2019-qy} experiment is
sequenced every ten generations, in the numerator for the covariance we use the
allele frequency changes between adjacent timepoints, which are ten generations
apart; \vb{thus the strong covariances between adjacent generations are not
  observable and do not contribute to the numerator of $G(t)$}.
  \vb{Additionally, selection also inflates the variance in allele frequency
    change per generation; however, this effect cannot be easily distinguished
  from drift.} \vb{For both these reasons, our measure $G(t)$ is strongly
conservative} (\vb{we demonstrate this through simulations in Supplementary
Material Section \ref{supp:tempblocks}}).  Still, we find a remarkably strong
signal. Greater than $20\%$ of total, genome-wide allele frequency change over
60 generations is the result of selection (Figure \ref{fig:figure-1} B). This
proportion of variance attributable to selection builds over time in Figure
\ref{fig:figure-1}B as the effects of linked selection are compounded over the
generations unlike genetic drift. Our G(t) starts to plateau to a constant
level as the covariances from earlier generations have decayed and so no longer
contribute as strongly (Figure \ref{fig:figure-1}).

@@

We observed that in the longest study  we analyzed \parencite{Barghi2019-qy},
some genome-wide temporal covariances become negative at future timepoints (see
the first two rows in Figure \ref{fig:figure-1} A). This shows that alleles
that were on average going up initially are later going down in frequency, i.e.
that the average direction of selection experienced by alleles has flipped.
This \vb{might} reflect either a change in the environment or the genetic
background, due to epistatic relationships among alleles altered by frequency
changes or recombination breaking up selective alleles; \vb{such reversals are
also theorized to occur after optima shifts (XXX)}.  Such reversals of
selective dynamics could be occurring at other timepoints but the signal of a
change in the direction of selection at particular loci may be washed out when
we calculate our genome-wide average temporal covariances. To address this
limitation, we calculated the distribution of the temporal covariances over
100kb windowed covariances (Figure \ref{fig:figure-3} shows these distributions
pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance estimate of each genomic window will be noisy, due to sampling and
genetic drift, and the neutral distribution of the covariance is complicated
due to linkage disequilibria, which can occur over long physical distances in
E\&R and selection studies \parencite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}. To
address this, we have developed a permutation-based procedure that constructs
an empirical \vb{neutral} null distribution by randomly flipping the sign of
the allele frequency changes in each genomic window (i.e. a single random sign
flip is applied to all loci in a window). This destroys the systematic
covariances created by linked selection and creates a sampling distribution of
the covariances spuriously created by neutral genetic drift while preserving
the complex dependencies between adjacent loci created by linkage
disequilibrium.  This empirical neutral null distribution is conservative in
the sense that the variances of the covariances are wider than expected under
drift alone as they include the effect of selection on the allele frequency
change within a time-interval, just not between time-intervals.  We see (Figure
\ref{fig:figure-3} A and B) that there are an empirical excess of windows with
positive covariances between close timepoints compared to the null distribution
(a heavier right tail), and that this then shifts to an excess of windows with
negative covariances between more distant timepoints (a heavier left tail)
